BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2016-12-28 Paper2/BOAST.tex
section boast use cases

Section~\ref{sec:laplace} presented the Laplace motivating example. From this
section we know that a number of optimizations can have an impact on the
performance of the kernel on the Mali architecture. But, what is the impact in
other architectures? And, are there additional optimizations that can impact
the performance?

@@

    \subsubsection{Performance Results}

@@

    \subsubsection{Laplace Conclusion}

@@

  \subsection{Creating an Auto-Tuned Convolution Library for BigDFT using
BOAST}

@@

Despite that the application is mainly written in Fortran, it currently includes
360 kLOC and 70 kLOC of Fortran and C languages, respectively, accounting for
more than 50\% of the code base.  It is a parallel application based on the
standards MPI~\cite{mpi} and OpenMP~\cite{openmp}.  It also supports
CUDA~\cite{cuda} and OpenCL~\cite{opencl}.  In the recent years this code has
been used for many scientific applications, and its development and user
consortium is continuously growing.  Massively parallel computations are
routinely executed with the BigDFT code, either in homogeneous or hybrid
architectures.  In 2009, the French Bull-Fourier award was attributed for the
implementation of the hybrid version of BigDFT~\cite{genovese2009density}.

@@

    \subsubsection{BigDFT}

@@

    \subsubsection{A Generic Convolution Library}

@@

    \subsubsection{Performance Report}

@@

  \subsection{Porting SPECFEM3D to OpenCL using BOAST}
  \label{subsec:specfem}

@@

    \subsubsection{SPECFEM3D}

@@

\subsubsection{Porting to OpenCL}

\paragraph{Porting kernels to BOAST}

Nvidia CUDA and OpenCL are based on the same programming model: a
massively parallel accelerator running in disjoint, non addressable
memory environment. Thanks to that proximity, we have been able to carry
out most of the porting task with only a limited knowledge about
SPECFEM3D internal physics.

@@

\paragraph{Porting run-time to OpenCL} The second part of the port
consisted in the translation of the CPU-side of the application, from
CUDA API to OpenCL API. Most of the functions of the interfaces are
very similar, with only naming-convention and data-structure
distinctions. Hence, it was clear that automatic rewriting tools
(namely \code{sed} regexp and \code{emacs-lisp} functions) could be
useful. To give an idea of the cost of a \emph{manual} rewriting, we
can count (in \# of OpenCL API function calls): 70 kernel ``function
calls'', 790 arguments to set, 230 memory transfers, 160 buffer
creations, and 270 releases.

@@

\paragraph{Debugging OpenCL Execution: \code{GPUTrace}}

With the help of \code{GPUTrace} (described in Section~\ref{sec:gputrace}), we
could confront CUDA and OpenCL execution traces with a graphical \code{diff}
tool, and spot the different porting mistakes: some parameters reversed,
offsets incorrectly applied, \etc{}.

@@

\paragraph{Evaluation} Our OpenCL/BOAST port of SPECFEM3D is now
merged in SPECFEM3D's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and N Intel
Xeon processors, we measured similar results between the original
CUDA version and our BOAST-CUDA version. With the same set of
optimization flags, BOAST CUDA and OpenCL version reported similar
execution time spans. The best execution speed was achieved with CUDA
version though (25\% higher than OpenCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed to OpenCL run-time, as of version
1.1. This parameter, in addition to specifying the work group size (which can
be done in OpenCL), also constrains the number of work group that must run in
parallel on a multiprocessor. This value is set to 7 in CUDA. This means that
the compiler must be very conservative on register usage in order to allow this
parallelism which allows better overlapping of communications end computations.
This functionality is not supported in OpenCL.

@@

\section{Related Work\label{sec:related}}

Code generation and auto-tuning techniques are not new. Nonetheless, with recent
developments in hardware, and the HPC landscape being as diverse as it is now,
there is a renewed interest in the field. This related work Section is split in
three parts, focusing first on Auto-tuning frameworks. Tools that provide a DSL
to describe computing kernels will then be presented. Last, optimization space
pruners and their ties to auto-tuning will be introduced.

>>>>

Section~\ref{sec:laplace} presented the Laplace motivating example. From this
section we know that a number of optimizations can have an impact on the
performance of the kernel on the Mali architecture. But, what is the impact in
other architectures? And, are there any additional optimizations that can impact
the performance?

@@

%%% PERFORMANCE RESULTS %%%
\subsubsection{Performance Results}

@@

%%% LAPLACE CONCLUSION %%%
\subsubsection{Laplace Conclusion}

@@

%%% CREATING AN AUTO-TUNED CONVOLUTION LIBRARY FOR BIGDFT USING BOAST %%%
\subsection{Creating an Auto-Tuned Convolution Library for BigDFT using BOAST}

@@

Despite that the application is mainly written in Fortran (360 kLOC of Fortran), it 
currently also includes 70 kLOC of C languages, accounting for
more than 50\% of the code base.  It is a parallel application based on the
standards MPI~\cite{mpi} and OpenMP~\cite{openmp}.  It also supports
CUDA~\cite{cuda} and OpenCL~\cite{opencl}.  In the recent years this code has
been used for many scientific applications, and its development and user
consortium is continuously growing.  Massively parallel computations are
routinely executed with the BigDFT code, either in homogeneous or hybrid
architectures.  In 2009, the French Bull-Fourier award was attributed for the
implementation of the hybrid version of BigDFT~\cite{genovese2009density}.

@@

%%% BIGDFT %%%
\subsubsection{BigDFT}

@@

%%% A GENERIC CONVOLUTION LIBRARY %%%
\subsubsection{A Generic Convolution Library}

@@

%%% PERFORMANCE REPORT %%%
\subsubsection{Performance Report}

@@

%%% Porting SPECFEM3D to OpenCL using BOAST %%%
\subsection{Porting SPECFEM3D to OpenCL using BOAST\label{subsec:specfem}}

@@

%%% SPECFEM3D %%%
\subsubsection{SPECFEM3D}

@@

%%% PORTING TO OPENCL %%%
\subsubsection{Porting to OpenCL}

%%%
\paragraph{Porting kernels to BOAST}  Nvidia CUDA and OpenCL are based on the same programming model: a
massively parallel accelerator running in disjoint, non addressable
memory environment. Thanks to that proximity, we have been able to carry
out most of the porting task with only a limited knowledge about
SPECFEM3D internal physics.

@@

%%%
\paragraph{Porting run-time to OpenCL} The second part of the port
consisted in the translation of the CPU-side of the application, from
CUDA API to OpenCL API. Most of the functions of the interfaces are
very similar, with only naming-convention and data-structure
distinctions. Hence, it was clear that automatic rewriting tools
(namely \code{sed} regexp and \code{emacs-lisp} functions) could be
useful. To give an idea of the cost of a \emph{manual} rewriting, we
can count (in \# of OpenCL API function calls): 70 kernel ``function
calls'', 790 arguments to set, 230 memory transfers, 160 buffer
creations, and 270 releases.

@@

%%%
\paragraph{Debugging OpenCL Execution: \code{GPUTrace}} 
With the help of \code{GPUTrace} (described in Section~\ref{sec:gputrace}), we
could confront CUDA and OpenCL execution traces with a graphical \code{diff}
tool, and spot the different porting mistakes: some parameters reversed,
offsets incorrectly applied, \etc{}.

@@

%%%
\paragraph{Evaluation} Our OpenCL/BOAST port of SPECFEM3D is now
merged in SPECFEM3D's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and N Intel
Xeon processors, we measured similar results between the original
CUDA version and our BOAST-CUDA version. With the same set of
optimization flags, BOAST CUDA and OpenCL version reported similar
execution time spans. The best execution speed was achieved with CUDA
version though (25\% higher than OpenCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed to OpenCL run-time, as of version
1.1. This parameter, in addition to specifying the work group size (which can
be done in OpenCL), also constrains the number of work group that must run in
parallel on a multiprocessor. This value is set to 7 in CUDA. This means that
the compiler must be very conservative on register usage in order to allow this
parallelism which allows better overlapping of communications end computations.
This functionality is not supported in OpenCL.

@@

%%% RELATED WORK %%%
\section{Related Work\label{sec:related}}

Code generation and auto-tuning techniques are not new. Nonetheless, with recent
developments in hardware, and the HPC landscape being as diverse as it is now,
there is a renewed interest in the field. This related work section is split in
three parts, focusing first on auto-tuning frameworks. Tools that provide a DSL
to describe computing kernels will then be presented. Last, optimization space
pruners and their ties to auto-tuning will be introduced.
